Aspects of Dealing with Imperfect Data in Temporal Databases

### 2012-03-16 book/sources/fuzzyTemporalDatabases.tex
To my opinion, the content is now perfect.

Data and information imperfections and techniques to represent them correctly in databases and queries are usually the focus of research in fuzzy databases. Proposals from this field may present an approach based somehow on fuzzy set theory~\cite{Zadeh65} or possibility theory~\cite{Dubois:Prade:1988:PossibilityTheory}, although other theories support information imperfections too. Comparably, many proposals concerning the introduction of data imperfections or information imperfections in temporal databases present approaches based somehow on fuzzy set theory~\cite{garrido2009},~\cite{gal01},~\cite{nagypal2003},~\cite{Billiet:Pons:Matthe:DeTre:Pons:2011:BipolarFuzzy} or possibility theory~\cite{Dubois89},~\cite{Pon11},~\cite{129895}, although proposals based on other theories exist~\cite{Dyreson1998},~\cite{Dekhtyar2001},~\cite{Qia09}. As possibility theory is usually seen as a theory of confidence, aimed at dealing with uncertainty, in some proposals, possibility theory is used specifically to handle uncertainty in temporal information. In fuzzy databases~\cite{Galindo2006}, uncertainty is usually expected to appear at storage level (in the database), whereas other types of imperfection, notably imprecision, are usually expected to appear in querying.

@@

In fact, to model the uncertainty related to a valid-time interval using possibility theory, the presented proposal relies on the concept of ill-known sets~\cite{Dubois88b}.

@@

The presented proposal relies on the concept of ill-known sets~\cite{Dubois88b}. To correctly explain the concept of ill-known sets, the concept of possibilistic variables should be introduced first. In the presented proposal, the definition of possibilistic variables of~\cite{Pon11} is followed. A \emph{possibilistic variable} is defined as follows~\cite{Pon11}.

@@

BASE ON IPMU 2

Consider a set $R$, which contains single values (and not collections of values). When a possibilistic variable $X_{v}$ is defined on such a set $R$, the unique value $X_{v}$ takes, which is (partially) unknown, is called an \emph{ill-known value} in this work \cite{Dubois88b}.

%The exact value a possibilistic variable takes, which is (partially) unknown, is called an \emph{ill-known value} in this work \cite{Dubois88b}.

When a possibilistic variable is defined on the powerset $\Pow(R)$ of some universe $R$, the unique value the variable takes will be a crisp set and its possibility distribution on the powerset $\Pow(R)$ will describe the possibility of each crisp subset of $R$ to be the value the variable takes. This exact value (a crisp set) the variable takes, is now called an \emph{ill-known set} \cite{Dubois88b}.

This work will deal with ill-known time intervals. These are ill-known intervals in time. Ill-known intervals are ill-known sets, defined and represented via a start and end point, which will be ill-known values. The elements of the set are the values between the start and end point. A closed ill-known interval with start point defined by possibilistic variable $X$ and end point by possibilistic variable $Y$ is noted here $\left[X, Y\right]$. The correspondences and transitions between the representations of ill-known sets, between the representations of ill-known intervals and between the representations of an ill-known set and an ill-known interval are part of the authors current research. Figure \ref{fig:interval} illustrates this.

@@

Several authors ~\cite{garrido2009} propose transformations in order to optimize the storage of such ill-known valid-time intervals, though recent research might seem to indicate some minor issues with respect to some of these transformations ~\cite{Pon11}. A comparison between the transformations in ~\cite{garrido2009} and the framework in~\cite{Pon11} is presented in~\cite{pon12}. Figure \ref{fig:convexhull} illustrates a transformation based on the `convex hull' approach REF.

%An ill-known time interval is usually represented by means of two possibility distributions one in each starting and ending points. Figure \ref{fig:interval} shows an ill-known valid-time interval. The starting and ending points are denoted by two fuzzy numbers, $X$ and $Y$, represented by means of two triangular possibility distributions (see the Appendix). For illustration, figure \ref{fig:convexhull} shows one of the transformations based in the convex hull and proposed in~\cite{garrido2009}.

\begin{figure}
\centering
\input{graphs/convexhull.tex}
\caption{Transformation based in the convex hull from the two ill-known points $X$ and $Y$. }
\label{fig:convexhull}
\end{figure}

In section \ref{subsec:querying-time-intervals}, a technique allowing the querying of fuzzy valid-time databases which contain ill-known valid-time intervals is presented and briefly explained. The technique relies on the framework for set evaluation presented in \cite{Pon11}.

\subsubsection{Ill-known Time Constraints}

\subsubsection{Querying Fuzzy Valid-Time Databases with Ill-Known Valid-Time Intervals}
\label{subsec:querying-time-intervals}

One of the main purposes of the existence of databases is to allow information retrieval. The standard query language for databases is SQL ~\cite{Mel93}. Several proposals to extend the SQL language for transaction-time databases~\cite{Sarda90}, valid-time databases~\cite{gad92} and bitemporal databases ~\cite{TSQL} exist and some authors have studied how to support temporal querying in standard SQL~\cite{Snodgrass98}.

In the following subsections, the query structure, the query evaluation method and the technique to meaningfully rank and aggregate the query results, proposed in this work, are presented.

@@

\paragraph{\textbf{Query Structure}}
In this work, a query $\tilde{Q}$ is made up of two separate constructs of user query demands.

@@

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}
\textbf{Query}
A query $\tilde{Q}$ in this work is given by:
\begin{equation}
\label{eq:query-specification}
\tilde{Q} = \left( Q^{time}, Q \right)
\end{equation}
Here $Q = \left \lbrace q_1,\ \cdots,\ q_n \right \rbrace$ contains all (possibly fuzzy) non-temporal user-defined query demands. These comprise all constraints unrelated to the valid-time indications in the queried relations. $Q^{time}$ denotes the temporal constraint specified by the user.
\end{definition}
\vspace{-10pt}
\end{svgraybox}
%%specify better this:
%\begin{svgraybox}
%\begin{definition}
%\label{def:query-constraint}
%\textbf{(query constraint)}
%$Q$ is the 
%A query constraint $q_a$ for an attribute $a$ with attribute domain $\mathcal{D}$, is a restriction in the subset of the values in the attribute domain:
%\begin{equation}
%\label{eq:query-constraint}
%q_a = \left \lbrace d \mid d \subseteq \mathcal{D} \right \rbrace
%\end{equation}
%\end{definition}

In this work, the query structure allows the user to specify a single temporal constraint.

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}
\textbf{Temporal Constraint $Q^{time}$}
In this work, $Q^{time}$ is defined by:
\begin{equation}
Q^{time} = \left( I,AR \right)
\end{equation}
where $I$ denotes a crisp time interval and $AR$ is one of the Allen relations.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

The interpretation is that, for a record with an ill-known valid-time interval $J$, the user demands that $I$ AR $J$ hold.

\paragraph{\textbf{Query evaluation}}
\label{subsubsec:query-evaluation}
Query satisfaction in a fuzzy relational database is a matter of degree. Typically, the query evaluation results in a \emph{satisfaction degree} $s$, which is typically in the unit interval. $s \in \left[ 0,1,\right]$. Here, a satisfaction degree of $0$ denotes a total dissatisfaction while a degree of $1$ denotes a total satisfaction.

@@

In the presented approach, for every record $r$, each part of $\tilde{Q}$ is evaluated independently:
\begin{itemize}
\item
The user preferences expressed in the non-temporal part, $Q$, are evaluated, resulting in a satisfaction degree denoted as $e_Q(r)$. The presented approach accepts any valid, sound method of calculating this evaluation, as long as the method is well-founded and $e_Q(r) \in \left[0,1\right]$.
\item
The evaluation of the temporal demand expressed in the temporal part, $Q^{time}$, depends on $AR$. A specific set of ill-known constraints\cite{Pon11} is considered depending on the Allen relation denoted by $AR$. The exact set chosen can be found in table \ref{tab:fuzzy-allen-relations}, for every possible value of $AR$. Then, using these formulas given in table \ref{tab:fuzzy-allen-relations}, both the possibility degree $\Pos_{Q^{time}}(r)$ and the necessity degree $\Nec_{Q^{time}}(r)$ are computed.
\end{itemize}

\textcolor{red}{TODO: maybe part on what these and mean? Maybe part on where the formulas in table \ref{tab:fuzzy-allen-relations} come from?}

\begin{table}[h]

\caption{Allen's relations used in the framework. Here, $I = \left[a, b\right]$ denotes a crisp time interval, $J = \left[X, Y\right]$ denotes an ill-known time interval, with $\pi_{X}$ and $\pi_{Y}$ the possibility distributions of $X$ and $Y$ respectively. The second column contains the corresponding formula to calculate the possibility that $I$ satisfies all constraints given by the Allen's relation.}

\centering
\begin{tabular}{|c|c|c|}
\hline
Allen Relation &  $\Pos\left(\text{I satisfies all constraints }C_{i}, i = 1,2,...\right)$ & $\Nec\left(C_{i}, i = 1,2,... \right)$ \\
\hline
I before J & $\sup_{a>w}\pi_X(w)$ & $\inf_{a \leq w} 1- \pi_X(w)$\\
\hline

\multirow{4}{*}
{I equal J} &  $\min ( \sup_{a \leq w}\pi_X(w),$  & $\min ( \inf_{a>w} 1-\pi_X(w),$\\
& $\pi_X(w),$ & $\inf_{w \in \left[a,b\right]} \pi_X(w), $ \\
 &  $\sup_{b \geq w}\pi_Y(w),$ & $\inf_{b>w} 1-\pi_Y(w),$\\
 & $\pi_Y(w))$ & $\inf_{w \in \left[a,b\right]} \pi_Y(w) )$ \\
\hline

\multirow{2}{*}
I meets J  & $\min (\sup_{a\geq w} \pi_X(w),$ & $\min (\inf_{a<w} 1 - \pi_X(w),$\\
& $\pi_X(w))$  & $ \inf_{w \in \left[a,b\right]} \pi_X(w) )$ \\
\hline

\multirow{3}{*}
{I overlaps J}  & $\min ( \sup_{b>w}\pi_Y(w), $ & $\min ( \inf_{b \geq w} 1 - \pi_Y(w),$\\
 & $\sup_{a \geq w}\pi_X(w),$& $1- \inf_{a>w}1-\pi_X(w),$\\
 & $\sup_{a \leq w}\pi_X(w))$ & $1- \inf_{a<w}1-\pi_X(w) )$ \\
\hline

\multirow{4}{*}
{I during J}  & $\max ( \min ( \sup_{a<w}\pi_X(w),,$ & $\max (\min (\inf_{a \leq w}1-\pi_X(w), $\\
 & $\sup_{b \geq w}\pi_Y(w)) $ & $ \inf_{b>w} 1- \pi_Y(w)),$ \\
 & $\min ( \sup_{a \leq w }\pi_X(w),$ & $\min ( \inf_{a<w}1-\pi_X(w),$\\
 & $\sup_{b>w}\pi_Y(w))$ & $\inf_{b\geq w}1-\pi_Y(w)))$ \\
\hline
\multirow{2}{*}
{I starts J} &  $\min( \sup_{a \leq w}\pi_X(w),$  & $\min (\inf_{a<w}1-\pi_X(w),$\\
& $\pi_X(w))$ & $inf_{w \in \left[a,b\right]} 1- \pi_X(w))$ \\
\hline
\multirow{2}{*}
{I finishes J} &  $\min ( \sup_{b \geq w} \pi_Y(w),$  & $\min(\inf_{b>w}1-\pi_Y(w),$\\
& $\pi_Y(w))$ & $\inf_{w \in [a,b]} 1- \pi_Y(w))$ \\
\hline

\end{tabular}
%
%\vspace{10pt}
%
%
%\vspace{-25pt}
\label{tab:fuzzy-allen-relations}
\end{table}

\subsubsection{Aggregation and Ranking} TO BE ANOTHER PART
To be able to present the most appropriate results to the user, for every record $r$, an aggregation method is used to aggregate $\Pos_{Q^{time}}(r)$ and $\Nec_{Q^{time}}(r)$ into a temporal record rank $e_{Q^{time}}(r)$ and after this, a convex combination combining $e_{Q^{time}}(r)$ and $e_{Q}(r)$ will provide the final record rank $e_{final}(r)$.

@@

>>>>

Data and information imperfections and techniques to represent them correctly in databases and queries are usually the focus of research in fuzzy databases. Proposals from this field may present an approach based somehow on fuzzy set theory~\cite{Zadeh65} or possibility theory~\cite{Dubois:Prade:1988:PossibilityTheory}, although other theories support information imperfections too. Comparably, many proposals concerning the introduction of data imperfections or information imperfections in temporal databases present approaches based somehow on fuzzy set theory~\cite{garrido2009},~\cite{gal01},~\cite{nagypal2003},~\cite{Billiet:Pons:Matthe:DeTre:Pons:2011:BipolarFuzzy} or possibility theory~\cite{Dubois89},~\cite{Pon11},~\cite{129895}, although proposals based on other theories exist~\cite{Dyreson1998},~\cite{Dekhtyar2001},~\cite{Qia09}. As possibility theory is usually seen as a theory of confidence, aimed at dealing with uncertainty, in some proposals, possibility theory is used specifically to handle uncertainty in temporal information. In fuzzy databases~\cite{Galindo2006}, uncertainty is usually expected to appear in the database content, whereas other types of imperfection, notably imprecision, are usually expected to appear in querying.

@@

In fact, to model the uncertainty related to a valid-time interval using possibility theory, the presented proposal introduces so-called ill-known time intervals, relying on the concept of ill-known sets~\cite{Dubois88b}.

@@

To represent valid time indications which might contain uncertainty, the presented proposal introduces the concept of ill-known valid time intervals, which relies on the concept of ill-known sets~\cite{Dubois88b}. To correctly explain the concept of ill-known sets, the concept of possibilistic variables should be introduced first. In the presented proposal, the definition of possibilistic variables of~\cite{Pon11} is followed. In~\cite{Pon11}, a \emph{possibilistic variable} is defined as follows.

@@

Now, consider a set $R$, which contains single values (and not collections of values). When a possibilistic variable $X_{v}$ is defined on such a set $R$, the unique value $X_{v}$ takes, which is (partially) unknown, is called an \emph{ill-known value} in this work \cite{Dubois88b}.

When a possibilistic variable is defined on the powerset $\Pow(R)$ of some universe $R$, the unique value the variable takes will be a crisp set and its possibility distribution on the powerset $\Pow(R)$ will describe the possibility of each crisp subset of $R$ to be the value the variable takes. This value (a crisp set) the variable takes, which is (partially) unknown, is now called an \emph{ill-known set} \cite{Dubois88b}.

Finally, consider a set $R$, which contains single values (and not collections of values) and its powerset $\Pow(R)$. Now consider a subset $\Pow_{I}(R)$ of $\Pow(R)$ and let this subset contain every element of $\Pow(R)$ that is an interval, but no other elements. When a possibilistic variable $X_{i}$ is defined on the subset $\Pow_{I}(R)$ of the powerset $\Pow(R)$ of some set $R$, the unique value $X_{i}$ takes will be a crisp interval and the possibility distribution $\pi_{X_{i}}$ of $X_{i}$ will be a possibility distribution on $\Pow_{I}(R)$. This $\pi_{X_{i}}$ will define the possibility of each value of $\Pow_{I}(R)$ (a value of $\Pow_{I}(R)$ is a crisp interval which is a subset of $R$) being the value $X_{i}$ takes. This exact value (a crisp interval) the variable takes, which is (partially) unknown, is called an \emph{ill-known interval} here.

The presented proposal will deal with ill-known time intervals. Ill-known time intervals are ill-known intervals in a time domain. In the presented proposal, an ill-known time interval will be defined and represented via its start and end point, which will be ill-known values. The elements of the ill-known time interval are the values between its start and end point, including the start and end points themselves\footnote{The presented proposal only deals with closed ill-known time intervals. Dealing with halfopen or open ill-known intervals is part of the current research of the authors.}. It should be clear that this approach to representing ill-known time intervals differs from the approach based on a single possibility distribution on a set $\Pow_{I}(R)$ of a set $R$. These approaches have a different behavior and can be used to describe different ill-known time intervals. The correspondences, interactions and transitions between these different representations of ill-known intervals and their interpretations are part of the authors current research.

In the presented proposal, a closed ill-known time interval with start point defined by possibilistic variable $X$ and end point by possibilistic variable $Y$ is noted $\left[X, Y\right]$. Figure \ref{fig:interval} shows a closed ill-known time interval. %The correspondences and transitions between the representations of ill-known sets, between the representations of ill-known intervals and between the representations of an ill-known set and an ill-known interval are part of the authors current research. Figure \ref{fig:interval} illustrates this.

@@

Several authors work with concepts very similar to these ill-known time intervals and some of them~\cite{garrido2009} propose transformations of the describing functions in order to optimize the storage of such ill-known valid time intervals, though recent research might seem to indicate some minor issues with respect to some of these transformations~\cite{Pon11}. A comparison between the transformations in~\cite{garrido2009} and the framework in~\cite{Pon11} is presented in~\cite{pon12}. Figure \ref{fig:convexhull} illustrates a transformation based on the `convex hull' approach~\cite{garrido2009}.

%An ill-known time interval is usually represented by means of two possibility distributions one in each starting and ending points. Figure \ref{fig:interval} shows an ill-known valid-time interval. The starting and ending points are denoted by two fuzzy numbers, $X$ and $Y$, represented by means of two triangular possibility distributions (see the Appendix). For illustration, figure \ref{fig:convexhull} shows one of the transformations based in the convex hull and proposed in~\cite{garrido2009}.
\begin{figure}
\centering
\input{graphs/convexhull.tex}
\caption{Transformation based in the convex hull from the two ill-known points $X$ and $Y$. }
\label{fig:convexhull}
\end{figure}

%In section \ref{subsec:querying-time-intervals}, a technique allowing the querying of fuzzy valid-time databases which contain ill-known valid-time intervals is presented and briefly explained. The technique relies on the framework for set evaluation presented in \cite{Pon11}.

In the presented proposal, ill-known valid time intervals will be used to represent valid time indications in a valid-time relation and to model the uncertainty these may contain.

To evaluate the temporal demands in queries issued by users to query a valid-time relation containing ill-known valid time intervals, the presented proposal introduces a technique based on the concept of ill-known time constraints, which is based on the concept of ill-known constraints as presented in~\cite{Pon11}. Both concepts are treated in subsection \ref{iktc}.

Before ill-known time constraints can be introduced, another notion related to possibilistic variables shoud be paid attention to. In fact, a specific application of possibilistic variables is obtained when the set under consideration is the set of boolean values, denoted $\mathbb{B}$ = $\{T,F\}$, where $T$ denotes `true' and $F$ denotes `false' \cite{Pon11}. Indeed, any boolean proposition $p$ takes exactly one value in $\mathbb{B}$. If the knowledge about which value this proposition $p$ takes, is given by a possibility distribution $\pi_p$, then proposition $p$ can be seen as a possibilistic variable. In the presented proposal, the interest lies with the case where the proposition holds (denoted $p$ = $T$) and the possibility and necessity that $p$ = $T$ demand most attention. In the following sections, the following notations are used, based on previous notations:

\begin{align}
\text{Possibility that $p$ = $T$:} \hspace{50pt} & Pos(p) = \pi_p(T) \\
\text{Necessity that $p$ = $T$:} \hspace{50pt} & Nec(p) = 1-\pi_p(F) 
\end{align}

\subsubsection{\label{iktc}Ill-known Time Constraints}
The presented proposal contains a technique for evaluating user queries used to query a valid time relation in which the valid time indications are represented by ill-known time intervals. Part of this query evaluation technique relies on the concept of ill-known time constraints, which is based on the concept of ill-known constraints as presented in~\cite{Pon11}. These concepts are presented below. Following~\cite{Pon11}, an \emph{ill-known constraint} is defined as follows.

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}
Given a universe $U$, an \emph{ill-known constraint} $C$ is specified by means of a binary relation $R \subseteq U^{2}$ and a fixed, ill-known value defined by its possibilistic variable $V$ on $U$, i.e.:
\begin{align}
C \triangleq (V,R) \nonumber
\end{align}
A set $A \subseteq U$ now satisfies this constraint $C$ if and only if:
\begin{align}
\forall a \in A : (V,a) \in R \nonumber
\end{align}
\end{definition}
\vspace{-10pt}
\end{svgraybox}

An example of an ill-known constraint is given by:
\begin{align}
C_{<} \triangleq (X, <) \nonumber
\end{align}
Some set $A$ then satisfies $C_{<}$ if 
\begin{align}
\forall a \in A : X <a \nonumber
\end{align}

The satisfaction of a constraint $C \triangleq (V,R)$ by a set $A$ is basically a Boolean matter (either the set satisfies the constraint or not) and can thus be seen as a boolean proposition, but due to the uncertainty inherent to the ill-known value $V$, it can be uncertain whether $C$ is satisfied by $A$ or not \cite{Pon11}. Based on the possibility distribution $\pi_{V}$ of $V$, the possibility and necessity that $A$ satisfies $C$ can be found. This proposition can thus be seen as a possibilistic variable on $\mathbb{B}$. The required possibility and necessity are calculated using the following formulas~\cite{Pon11}.

\begin{align}
Pos(A\text{ satisfies }C) & = \min_{a \in A}\left(\sup_{(w,a) \in R}\pi_{V}(w))\right) \label{eq:pos}\\
Nec(A\text{ satisfies }C) & = \min_{a \in A}\left(\inf_{(w,a) \notin R} 1-\pi_{V}(w)\right) \label{eq:nec}
\end{align}

Now, to calculate the possibility or necessity of a set $A$ satisfying multiple constraints, the $\min$ t-norm operator is used to express a conjunction of constraints. For example:

\begin{align}
Pos((A\text{ satisfies }C_{1}) \text{ and } (A\text{ satisfies }C_{2})) & = \nonumber\\
 \min_{a \in A}(Pos(A\text{ satisfies }C_{1}), Pos(A\text{ satisfies }C_{2})) \nonumber \\
Nec((A\text{ satisfies }C_{1}) \text{ and } (A\text{ satisfies }C_{2})) & = \nonumber\\
 \min_{a \in A}(Nec(A\text{ satisfies }C_{1}), Nec(A\text{ satisfies }C_{2})) \nonumber
\end{align}

Accordingly, the $\max$ t-conorm operator is used to express a disjunction of constraints. For example:

\begin{align}
Pos((A\text{ satisfies }C_{1}) \text{ or } (A\text{ satisfies }C_{2})) & = \nonumber\\
 \max_{a \in A}(Pos(A\text{ satisfies }C_{1}), Pos(A\text{ satisfies }C_{2})) \nonumber \\
Nec((A\text{ satisfies }C_{1}) \text{ or } (A\text{ satisfies }C_{2})) & = \nonumber\\
 \max_{a \in A}(Nec(A\text{ satisfies }C_{1}), Nec(A\text{ satisfies }C_{2})) \nonumber
\end{align}

%Negation?

In the presented proposal, \emph{ill-known time constraints} are considered, which are ill-known constraints of which the considered universe is a time domain.

In the next subsection, the core of the presented proposal is described.

\subsubsection{Querying Valid-time Relations containing Ill-known Valid-time Intervals}
\label{subsec:querying-time-intervals}

One of the main purposes of the existence of (relational) databases is to allow information retrieval. The standard query language for (relational) databases is SQL~\cite{Mel93}, but several proposals to extend the SQL language for transaction-time databases~\cite{Sarda90}, valid-time databases~\cite{gad92} and bitemporal databases~\cite{TSQL} exist and some authors have studied how to support temporal querying in standard SQL~\cite{Snodgrass98}.

As mentioned before, the presented proposal deals with querying a valid time relation. In this subsection, the core of the presented proposal is described. First the particular structure of the relation is described, along with the nature and structure of its supposed contents. Next, the particular query structure is presented. Finally, the particular method for evaluating queries and for ranking the result records are presented.

%In the following subsections, the nature and structure of the relation content, the query structure, the query evaluation method and the technique to meaningfully rank and aggregate the query results, proposed in this work, are presented.

@@

\paragraph{\textbf{Relation Structure}}
In the presented proposal, a valid-time relation is considered, in which every record contains just one valid-time indication. This valid-time indication is represented by a closed ill-known time interval, to allow uncertainty in the valid-time information. As explained above, the ill-known time intervals used here will be defined and represented via their start and end points, which are ill-known values in the valid-time domain.

\paragraph{\textbf{Query Structure}}
In the presented proposal, a query consists of two separate constructs of user demands.

@@

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}
\textbf{Query}\\
A query $\tilde{Q}$ is given by:
\begin{equation}
\label{eq:query-specification}
\tilde{Q} = \left( Q^{time}, Q \right)
\end{equation}

Here, $Q$ denotes the construct of all (possibly fuzzy) non-temporal user query demands. These comprise all constraints and demands unrelated to valid-time and thus unrelated to the valid-time indications in the queried relation. $Q^{time}$ denotes the temporal demand specified by the user.

%Here $Q = \left \lbrace q_1,\ \cdots,\ q_n \right \rbrace$ contains all (possibly fuzzy) non-temporal user-defined query demands. These comprise all constraints unrelated to the valid-time indications in the queried relations. $Q^{time}$ denotes the temporal constraint specified by the user.
\end{definition}
\vspace{-10pt}
\end{svgraybox}
%%specify better this:
%\begin{svgraybox}
%\begin{definition}
%\label{def:query-constraint}
%\textbf{(query constraint)}
%$Q$ is the 
%A query constraint $q_a$ for an attribute $a$ with attribute domain $\mathcal{D}$, is a restriction in the subset of the values in the attribute domain:
%\begin{equation}
%\label{eq:query-constraint}
%q_a = \left \lbrace d \mid d \subseteq \mathcal{D} \right \rbrace
%\end{equation}
%\end{definition}

The presented query structure allows the user to specify a single temporal demand, denoted by $Q^{time}$.

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}
\textbf{Temporal demand}\\
A temporal demand $Q^{time}$ is defined by:
\begin{equation}
Q^{time} = \left( I,AR \right)
\end{equation}
Here, $I$ denotes a crisp time interval (which can be specified in any way required) and $AR$ denotes one of the Allen relations (cf. section~\ref{allen-crisp}).
\end{definition}
\vspace{-10pt}
\end{svgraybox}

The interpretation of such a temporal demand $Q^{time} = \left( I,AR \right)$ is that, for a record with an ill-known valid-time interval $J$, the user demands that $I$ AR $J$ holds.

\paragraph{\textbf{Query evaluation}}
\label{subsubsec:query-evaluation}
Query satisfaction in a fuzzy relational database is usually a matter of degree. Typically, the evaluation of the query demands for a record results in a \emph{satisfaction degree} $s$, which is typically in the unit interval, i.e. $s \in \left[ 0,1\right]$. This satisfaction degree then models the extent to which the record satisfies the query demands. As such, a satisfaction degree of $0$ indicates total dissatisfaction while a degree of $1$ indicates total satisfaction and every level of satisfaction between total satisfaction and total dissatisfaction is indicated by a satisfaction degre between $0$ and $1$.

@@

In the presented approach, for every record $r$, each part of a query $\tilde{Q} = \left( Q^{time}, Q \right)$ is evaluated independently:
\begin{itemize}
\item
The user preferences expressed in the non-temporal part $Q$, are evaluated, resulting in a satisfaction degree denoted $e_Q(r)$. The presented approach accepts any valid, sound method of calculating this evaluation, as long as the method is well-founded and $e_Q(r) \in \left[0,1\right]$.
\item
The evaluation of the temporal demand expressed in the temporal part, $Q^{time} = \left( I,AR \right)$, depends on $AR$. A specific construct of ill-known constraints (cf. section~\ref{iktc}) is considered, depending on the Allen relation denoted by $AR$. The exact construct of constraints is an instantiation based on the formulas which can be found in table \ref{table:const}, for every possible value of $AR$. The form and capacity of these constraints are based on~\cite{Pon11}. Then, using equations \eqref{eq:pos} and \eqref{eq:nec}, the exact formulas to calculate the possibility $\Pos_{Q^{time}}(r)$ and the necessity $\Nec_{Q^{time}}(r)$ that record $r$ satisfies this construct of ill-known time constraints are derived from this construct of constraints. As mentioned, $\Pos_{Q^{time}}(r)$ and $\Nec_{Q^{time}}(r)$ denote the possibility, respectively the necessity that the considered record $r$ satisfies the construct of constraints corresponding to the temporal demand $Q^{time}$ and thus the possibility, respectively the necessity that $r$ satisfies $Q^{time}$.

%The exact set of constraints is an instantiation based on the formulas which can be found in table \ref{tab:fuzzy-allen-relations}, for every possible value of $AR$. Then, using these formulas given in table \ref{tab:fuzzy-allen-relations}, both the possibility degree $\Pos_{Q^{time}}(r)$ and the necessity degree $\Nec_{Q^{time}}(r)$ are computed.
\end{itemize}

%\textcolor{red}{TODO: maybe part on what these pos and nec mean? Maybe part on where the formulas in table \ref{tab:fuzzy-allen-relations} come from?}

\begin{table}[h]
\caption{Constructs of constraints related to their respective Allen relations, as used in the presented work. In this table, the ill-known time interval $J = \left[X, Y\right]$ in a record $r$ has a start point described by possibilistic variable $X$ and an end point described by possibilistic variable $Y$. The crisp time interval in the user's temporal demand is denoted $I$.}

\centering
\begin{tabular}{|c|c|}
\hline
Allen Relation & Construct of constraints\\
\hline
I before J & $C_1 \triangleq \left(<,X\right)$ \\
\hline
I equal J & $\left(C_1 \triangleq \left(\geq,X\right)\right) \wedge \neg \left(C_2 \triangleq \left(\neq,X\right) \right) \wedge \left(C_3 \triangleq \left(\leq,Y\right) \right) \wedge \neg \left(C_4 \triangleq \left(\neq,Y\right)\right)$ \\
\hline
I meets J & $\left(C_1 \triangleq \left(\leq,X\right)\right) \wedge \neg \left(C_2 \triangleq \left(\neq,X\right) \right)$ \\
\hline
I overlaps J & $\left(C_1 \triangleq \left(<,Y\right)\right) \wedge \neg \left(C_2 \triangleq \left(\leq,X\right) \right) \wedge \neg \left(C_3 \triangleq \left(\geq,X\right) \right)$ \\
\hline
I during J & $\left( \left (C_1 \triangleq \left(>,X\right) \right) \wedge \left(C_2 \triangleq \left(\leq,Y\right) \right) \right) \vee \left( \left(C_3 \triangleq \left(\geq,X\right) \right) \wedge  \left(C_4 \triangleq \left(<,Y\right) \right) \right)$ \\
\hline
I starts J & $\left(C_1 \triangleq \left(\geq,X\right) \right) \wedge \neg \left(C_2\triangleq \left(\neq,X\right)\right)$ \\
\hline
I finishes J & $\left(C_1 \triangleq \left(\leq,Y\right) \right) \wedge \neg \left(C_2\triangleq \left(\neq,Y\right)\right)$ \\
\hline
\end{tabular}
\label{table:const}
\end{table}

%\begin{table}[h]

%\caption{Allen's relations used in the framework. Here, $I = \left[a, b\right]$ denotes a crisp time interval, $J = \left[X, Y\right]$ denotes an ill-known time interval, with $\pi_{X}$ and $\pi_{Y}$ the possibility distributions of $X$ and $Y$ respectively. The second column contains the corresponding formula to calculate the possibility that $I$ satisfies all constraints given by the Allen's relation.}

%\centering
%\begin{tabular}{|c|c|c|}
%\hline
%Allen Relation &  $\Pos\left(\text{I satisfies all constraints }C_{i}, i = 1,2,...\right)$ & $\Nec\left(C_{i}, i = 1,2,... \right)$ \\
%\hline
%I before J & $\sup_{a>w}\pi_X(w)$ & $\inf_{a \leq w} 1- \pi_X(w)$\\
%\hline

%\multirow{4}{*}
%{I equal J} &  $\min ( \sup_{a \leq w}\pi_X(w),$  & $\min ( \inf_{a>w} 1-\pi_X(w),$\\
%& $\pi_X(w),$ & $\inf_{w \in \left[a,b\right]} \pi_X(w), $ \\
% &  $\sup_{b \geq w}\pi_Y(w),$ & $\inf_{b>w} 1-\pi_Y(w),$\\
% & $\pi_Y(w))$ & $\inf_{w \in \left[a,b\right]} \pi_Y(w) )$ \\
%\hline

%\multirow{2}{*}
%I meets J  & $\min (\sup_{a\geq w} \pi_X(w),$ & $\min (\inf_{a<w} 1 - \pi_X(w),$\\
%& $\pi_X(w))$  & $ \inf_{w \in \left[a,b\right]} \pi_X(w) )$ \\
%\hline

%\multirow{3}{*}
%{I overlaps J}  & $\min ( \sup_{b>w}\pi_Y(w), $ & $\min ( \inf_{b \geq w} 1 - \pi_Y(w),$\\
% & $\sup_{a \geq w}\pi_X(w),$& $1- \inf_{a>w}1-\pi_X(w),$\\
% & $\sup_{a \leq w}\pi_X(w))$ & $1- \inf_{a<w}1-\pi_X(w) )$ \\
%\hline

%\multirow{4}{*}
%{I during J}  & $\max ( \min ( \sup_{a<w}\pi_X(w),,$ & $\max (\min (\inf_{a \leq w}1-\pi_X(w), $\\
% & $\sup_{b \geq w}\pi_Y(w)) $ & $ \inf_{b>w} 1- \pi_Y(w)),$ \\
% & $\min ( \sup_{a \leq w }\pi_X(w),$ & $\min ( \inf_{a<w}1-\pi_X(w),$\\
% & $\sup_{b>w}\pi_Y(w))$ & $\inf_{b\geq w}1-\pi_Y(w)))$ \\
%\hline
%\multirow{2}{*}
%{I starts J} &  $\min( \sup_{a \leq w}\pi_X(w),$  & $\min (\inf_{a<w}1-\pi_X(w),$\\
%& $\pi_X(w))$ & $inf_{w \in \left[a,b\right]} 1- \pi_X(w))$ \\
%\hline
%\multirow{2}{*}
%{I finishes J} &  $\min ( \sup_{b \geq w} \pi_Y(w),$  & $\min(\inf_{b>w}1-\pi_Y(w),$\\
%& $\pi_Y(w))$ & $\inf_{w \in [a,b]} 1- \pi_Y(w))$ \\
%\hline

%\end{tabular}
%
%\vspace{10pt}
%
%
%\vspace{-25pt}
%\label{tab:fuzzy-allen-relations}
%\end{table}

%Thus, the evaluation of a query $\tilde{Q}$ for a record $r$ thus results in a satisfaction degree related to the non-temporal query demands and a couple of possibility and necessity related to the temporal demand. To be able to present the most appropriate results to the user most prominently, these evaluation results are aggregated and combined to result in a record rank.

\paragraph{\textbf{Aggregation and Ranking}} 
In this subsection, the notations used in the previous subsection are followed. To be able to present the most appropriate results to the user most prominently, for every record $r$, an aggregation method is used to aggregate $\Pos_{Q^{time}}(r)$ and $\Nec_{Q^{time}}(r)$ into a temporal record rank $e_{Q^{time}}(r)$ and after this, a convex combination combining $e_{Q^{time}}(r)$ and $e_{Q}(r)$ will provide the final record rank $e_{final}(r)$.

@@

In the next subsection, some main concepts and issues concerning bipolarity in the context of temporal databases are presented and discussed.
